%% ================================================================
%%  SECTION 6 — Judicial Outcome Simulation
%% ================================================================
\section{Judicial Outcome Simulation}

Of the five axes we survey, this one has the longest research history. The methods span decades---from bag-of-words classifiers fed into SVMs, through transformer-based models, all the way to today's billion-parameter language models fine-tuned on legal corpora. The task itself is easy to state: given the text of a case, predict whether it will be accepted, rejected, or partially accepted, and ideally explain why. How we evaluate that task, though, is in flux~\cite{nigam-etal-2024-rethinking}. We have come to believe that accuracy numbers, taken alone, are nearly meaningless in a legal context. A judge does not care that your model is 87\% accurate if it cannot articulate why a particular case should be dismissed.

\subsection{Classical and Early Neural Approaches}

The first generation of judgment prediction systems relied on hand-crafted features---bag-of-words vectors, TF-IDF scores, manually defined statutory indicators---piped into SVMs or logistic regression. They worked, sort of. Accuracy on small curated datasets was modest, and generalisation across court levels, case types, and time periods was poor. Moving to CNNs and RNNs eliminated the manual feature bottleneck by letting models ingest raw text directly, but Indian judgments are long. Really long. Standard recurrent architectures choke on documents that routinely exceed their input capacity, and no amount of gradient clipping fixes the fundamental context limitation. LegalBERT was the field's first serious attempt to address the mismatch between general-purpose language models and legal text by pre-training on legal corpora specifically~\cite{chalkidis-etal-2020-legalbert}. We would argue, in hindsight, that its real contribution was not the accuracy gains themselves but the demonstration that domain shift in legal NLP is severe enough to warrant dedicated pre-training.

\subsection{Transformer-Based Prediction}

Transformers changed what was possible. LegalBERT~\cite{chalkidis-etal-2020-legalbert} and InLegalBERT~\cite{paul-etal-2023-inlegalbert} brought domain-specific pre-training into the picture, and the accuracy gains were immediate. ILDC made systematic evaluation possible for the Indian context, offering over 35,000 Supreme Court cases annotated with binary outcome labels and a multi-sentence explanation layer~\cite{malik-etal-2021-ildc}. But the 512-token context window remains a stubborn constraint. Indian judgments blow past it routinely. Hierarchical encoders, overlapping chunk strategies, Longformer-style extended attention---all of these have been tried, and all of them introduce their own training and inference complexity while still imposing a ceiling on how much of the judgment actually informs the final prediction. We suspect that this ceiling is more damaging than current evaluations reveal, because most benchmarks happen to include relatively shorter Supreme Court orders rather than the sprawling high court judgments where the truncation problem bites hardest.

\subsection{Explainability as a First-Class Requirement}

A model that outputs ``dismissed'' and nothing else is useless to a lawyer. Explainability in legal prediction is not a nice-to-have; it is a functional requirement. Without it, you cannot audit for bias, you cannot identify error patterns, and you certainly cannot build trust with the judiciary. PredEx tackles this head-on with a corpus of over 15,000 Indian court cases where natural language explanations are aligned to prediction outcomes~\cite{nigam-etal-2024-predex}. Crucially, these are not post-hoc rationalisations that a model generated to justify its own output. They are written by legal professionals and reflect the actual reasoning behind each judgment---a distinction that matters more than most papers acknowledge. Attention-based explanation proxies, which dominated the field for a while, have been shown repeatedly to be unreliable indicators of what the model actually learned. PredEx sidesteps this problem entirely by grounding explainability in expert annotation rather than model introspection. We consider it the current gold standard for explanation evaluation in Indian legal NLP, though whether expert annotations themselves capture the full richness of judicial reasoning is a question that deserves more scrutiny than it has received.

\subsection{Large Language Models and the Scale Frontier}

Scale has arrived. NyayaAnumana assembles over 700,000 Indian court proceedings---district courts, high courts, Supreme Court---and ships with INLegalLlama, a domain-specialised LLaMA variant evaluated across this corpus~\cite{nigam-etal-2025-nyayaanumana}. At this scale, models begin to absorb the distributional texture of judicial reasoning: how different court levels frame issues, which statutory domains skew toward acceptance, what language patterns correlate with dismissal. The gains are real.

But so is the evaluation problem. Standard judgment prediction benchmarks select cases retrospectively, which means they are biased toward clear-cut outcomes. Genuinely contested disputes---the ones where prediction would actually be useful---are underrepresented~\cite{nigam-etal-2024-rethinking}. We think this benchmark bias is the single most underappreciated methodological issue in legal judgment prediction today. Retrieval-augmented generation currently produces the best results, because retrieved precedents supply contextual grounding that parametric knowledge alone cannot provide~\cite{nigam-etal-2025-nyayarag}. The implication loops back to Section~III: prediction quality is bottlenecked by retrieval quality, which makes the continued isolation of these two tasks all the more frustrating.

\subsection{Remaining Limitations}

We see three concrete problems that the field has not yet solved. The first is isolation. Prediction models train and evaluate without any connection to the retrieval and structuring systems we reviewed in Sections~III and~IV. Structured case facts, retrieved precedents---none of it feeds into the predictor. The pipeline is disconnected at exactly the point where connection would matter most.

The second is data bias toward the Supreme Court. Most benchmarks draw from Supreme Court judgments, but district courts and high courts handle the overwhelming majority of India's fifty-million-case backlog. These lower courts write differently, handle different case types, and produce different outcome distributions. Evaluating only on Supreme Court data and claiming generalisable results is, to put it bluntly, misleading.

Third, long documents. Even LLMs with extended context windows---128k tokens and beyond---have not been rigorously tested on Indian judgments that run past 50,000 words. We do not know how they degrade, because nobody has checked. And cutting across all three limitations is a quieter methodological concern: the community defaults to accuracy as its primary metric, sidelining macro-averaged F1, calibration error, and per-class recall without adequate justification~\cite{medvedeva-etal-2020-echr}. Accuracy on an imbalanced dataset tells you almost nothing about a model's real-world reliability.

%% ================================================================
%%  SECTION 7 — What-If Analysis
%% ================================================================
\section{What-If Analysis}

This axis has almost no literature behind it. It is also, in our view, the one that matters most to actual lawyers. Prediction asks what the outcome is. Explanation asks why. What-if analysis asks a different question entirely: what would need to change---in the facts, the statutory framing, or the precedent basis---for the outcome to flip? That distinction sounds subtle, but its practical consequences are enormous. The broader XAI literature recognises counterfactual explanation as one of the most actionable forms of model interpretability, precisely because it converts model behaviour into concrete guidance that a user can act on~\cite{mersha-etal-2024-xai}. We would go further and argue that for legal practitioners specifically, counterfactual reasoning is not just one form of explainability among many---it is the form that maps most directly onto how lawyers already think about case strategy.

\subsection{Counterfactual Explanation in NLP}

The technical formulation is straightforward. Given a text input and a model prediction, generate a minimally modified version of the input that flips the output. Which words, phrases, or factual claims, if changed, would produce a different label? The XAI literature offers several ways to approach this---gradient-based saliency, attention weight analysis, perturbation-based methods~\cite{mersha-etal-2024-xai}---but a crucial distinction often gets lost in the taxonomy. Feature importance tells you what mattered for the current prediction. Counterfactual explanation tells you what would need to be different. These sound related; they are not equivalent. A feature importance map might highlight the phrase ``dowry demand'' as decisive in a conviction prediction, but that is backward-looking. What a defence lawyer actually needs to know is: if the witness testimony were framed differently, or if a particular precedent were introduced, would the prediction change? That forward-looking, actionable framing is what makes counterfactual reasoning uniquely suited to legal practice.

\subsection{XAI in Legal Decision-Making}

Explainability in legal AI is not just a technical problem. In jurisdictions with transparency obligations, affected parties have a right to understand the basis of decisions that touch their legal interests~\cite{deakin-etal-2023-xai-law}. That normative demand creates an uncomfortable tension with the current state of the art. The best-performing prediction models tend to be the most opaque. The most interpretable models tend to be the least accurate. LIME and SHAP have both been applied to legal classifiers, but neither was designed for counterfactual reasoning; they answer the ``what mattered'' question, not ``what would need to change.'' PredEx's expert-annotated explanations push toward legally meaningful interpretability~\cite{nigam-etal-2024-predex}, but they explain observed predictions---they do not explore hypothetical alternatives. The gap between explaining what happened and exploring what could happen is, we believe, the most consequential open challenge in legal XAI. Until systems can answer ``what if,'' their explanations will remain descriptive rather than prescriptive.

\subsection{The Indian Legal Gap and Practical Motivation}

As of this writing, nobody has applied counterfactual what-if analysis to Indian legal AI. The closest existing work involves sensitivity analyses on European legal prediction models, and those have not been adapted to Indian statutory structures, institutional hierarchies, or linguistic patterns. The gap is complete.

The practical scenarios are easy to imagine. Could a lawyer test whether reframing a charge under a different IPC section would change the predicted outcome? Could a judge explore whether citing an additional precedent would shift the analysis? These are not hypothetical questions---they describe the kind of reasoning that practitioners already perform manually, case by case, at enormous cost in time and cognitive effort. In jurisdictions with limited annotated data and constrained computational resources, which describes India precisely, what-if analysis is especially attractive because it permits structured exploration of the decision space without model retraining or new annotations~\cite{medvedeva-etal-2020-echr}. It is, in effect, a low-cost amplifier for the prediction and argument graph capabilities we surveyed in Sections~V and~VI.

We consider counterfactual reasoning the capability most likely to close the loop between retrieval, structured analysis, prediction, and interactive deliberation~\cite{mersha-etal-2024-xai, deakin-etal-2023-xai-law}. It is also, candidly, the capability for which the Indian NLP community has the farthest to go.

