%% ================================================================
%%  SECTION 10 — Gaps and Contribution
%% ================================================================
\section{Gaps and Contribution}

The central finding of this survey is that no existing Indian legal AI system integrates precedent retrieval, case fact structuring, argument graph construction, outcome prediction, and what-if analysis into a unified pipeline. Each of these five capabilities has been studied in isolation, and the connections that would link them into a coherent reasoning chain remain unimplemented in any published system. Retrieved precedents are not fed into structured fact analysis modules. Structured facts and entity annotations are not consumed by argument graph constructors. Argument graphs do not inform prediction models. Prediction outputs do not serve as the basis for counterfactual exploration. The closest existing work to pipeline integration couples retrieval with generation~\cite{nigam-etal-2025-nyayarag} or retrieval with case-based reasoning~\cite{wiratunga-etal-2024-cbrrag}, but even these systems leave structuring, argument modelling, and what-if analysis entirely outside their scope.

Three task-level gaps stand out as particularly consequential. First, argument graph construction for Indian legal judgments has no published dataset, trained model, or evaluation benchmark, making it the field's most significant structural blind spot. Second, counterfactual what-if analysis has not been attempted in the Indian legal context, despite being the capability most directly useful to practising lawyers preparing or challenging legal arguments. Third, evaluation across the surveyed tasks is systematically skewed toward Supreme Court data~\cite{malik-etal-2021-ildc, nigam-etal-2024-predex}, leaving district and high court levels without reliable benchmarks despite their collective responsibility for the overwhelming majority of India's pending cases. The annotation scope of existing shared tasks such as LegalEval further reflects this imbalance~\cite{modi-etal-2023-semeval}.

Three data-level gaps compound these task-level deficiencies. First, no annotated dataset captures argument-level structure within Indian judgments, linking claims to their supporting evidence, statutory warrants, and cited precedents. Second, no dataset addresses the multilingual character of Indian court proceedings, where Hindi and regional language passages embedded within English judgments remain unprocessed by all existing models. Third, the large-scale datasets that enable LLM fine-tuning, such as NyayaAnumana~\cite{nigam-etal-2025-nyayaanumana}, do not provide the expert-level explanation annotations available in smaller datasets such as PredEx~\cite{nigam-etal-2024-predex}. No existing resource combines both scale and annotation depth simultaneously.

The contribution of this survey is precisely circumscribed. It provides the first structured review of Indian legal NLP organised around five functional capability axes, maps the existing literature to each axis, identifies the datasets and evaluation methods associated with each, and produces a consolidated synthesis of what remains unsolved. The primary finding is that the field has matured sufficiently in individual capabilities to make integration a realistic next step, and that the absence of such integration is now the binding constraint on further progress. This survey serves as a reference point for researchers and practitioners seeking to build toward a unified Indian legal AI architecture.

%% ================================================================
%%  SECTION 11 — Ethical Considerations
%% ================================================================
\section{Ethical Considerations}

Any AI system trained on historical Indian court judgments inherits the biases embedded in those judgments. Indian legal history contains well-documented disparities in outcomes correlated with gender, caste, economic status, and geographic region. A prediction model trained to replicate historical outcome distributions will replicate historical disparities unless active debiasing measures are applied during training and evaluation. This risk is not unique to India; bias amplification has been identified as a general problem in legal prediction models across jurisdictions~\cite{medvedeva-etal-2020-echr}. The risk is, however, compounded in any system that influences real legal decisions, because model outputs may be treated as authoritative even when they reflect statistical artefacts rather than legal reasoning. Bias auditing across demographic and geographic subgroups must therefore be a mandatory component of any deployed Indian legal AI system, not an optional post-deployment analysis.

In legal contexts, explainability is not a technical convenience but a procedural requirement. Parties to litigation have a right to understand the basis on which decisions affecting their legal interests are made, and this right extends to any AI-assisted component of the decision-making process~\cite{deakin-etal-2023-xai-law}. A specific tension arises between the opacity of high-performing large language models and the transparency obligations that govern judicial proceedings. Generated text that is fluent and well-structured may nonetheless be legally unsound, and the distinction between fluency and soundness is not detectable by automated metrics alone. The move toward expert-annotated explanations represented by PredEx~\cite{nigam-etal-2024-predex} is the correct methodological direction, as it grounds the evaluation of explanations in domain expertise rather than surface-level text similarity. Fluency of generated text must not be mistaken for legal soundness.

A clear and unambiguous distinction must be drawn between a legal AI system as a decision-support tool and one as an autonomous adjudicator. The systems surveyed in this paper are appropriately characterised as the former: tools that organise information, surface relevant precedents, structure arguments, and estimate likely outcomes to assist human legal professionals in their reasoning. They are not, and should not be positioned as, replacements for judicial reasoning. Even the most capable current systems operate under significant limitations in coverage, robustness, and interpretability that preclude autonomous deployment~\cite{nigam-etal-2025-nyayaanumana, mersha-etal-2024-xai}. The responsible development of Indian legal AI requires sustained collaboration among NLP researchers, legal professionals, and policymakers, ensuring that technical capability is matched by institutional accountability and that the augmentation of legal reasoning does not erode the procedural safeguards on which the rule of law depends.

\bibliography{survey_references}
\end{document}
