\section*{Ethics Statement}
Ethical conduct was a cornerstone in our research, especially considering the sensitive nature of the data and the methodologies involved. In collecting and annotating the \texttt{PredEx} dataset, we ensured that the law students involved in the annotation process were treated fairly and compensated appropriately. Their consent was obtained for all participation, and while they made significant contributions to the dataset, they are not listed as authors of this paper. This distinction is made to acknowledge their contribution while also maintaining the academic integrity of the publication process.

Significantly, the senior legal expert who played a pivotal role in mentoring the annotation process, as well as providing guidance on the Likert rating system and evaluating the generated explanations for both the PredEx and ILDC datasets, is credited as one of the authors of this paper. This inclusion reflects the expert's substantial intellectual contribution to the research, in line with ethical norms and authorship guidelines in academic publishing.

Moreover, for the computational resources used in this study, we adhered to ethical standards by duly paying the subscription fees for Google Colab Pro. This payment ensured legitimate access to the necessary paid cloud services, which were instrumental in the development and testing of our AI models. We believe in supporting the services and platforms that enable research like ours, and this includes the responsible financial support of technology providers.

In summary, our approach to ethics encompassed not only the respectful and fair treatment of all individuals involved but also the adherence to legal and financial obligations. This comprehensive ethical stance underscores our commitment to conducting research that is not only innovative and impactful but also responsible and respectful of all parties involved.
