\section{Methodology}
This section outlines the methodology employed in our research for the tasks of Judgment Prediction and Judgment Prediction with Explanation.

\subsection{Judgment Prediction}
\subsubsection{Language Model based}
In our approach, we utilized several language models including InCaseLaw, InLegalBERT \cite{paul-2022-pretraining}, XLNet (large) \cite{yang2019xlnet}, and Roberta (large) \cite{liu2019roberta} as baselines for binary classification. Due to the length constraints of complete judgments, which exceed the token capacity of these models, we adopted a chunking strategy. Each document was divided into 512-token chunks using a moving window approach with a 100-token overlap to preserve textual context. For model training, we used a batch size of 16, the Adam optimizer \cite{kingma2014adam}, and a learning rate of 2e-6. The training was conducted over 5 epochs on the PredEx train dataset. The remaining hyperparameters were set to their default values as provided by the HuggingFace library.

\subsubsection{Large Language Model based}
For utilizing LLMs in prediction, we employed two strategies: one involving prediction instructions only, and the other combining prediction with explanation instructions. Various models like Zephyr \cite{tunstall2023zephyr}, Gemini 1.0 Pro \cite{team2023gemini}, Llama-2-7B \cite{touvron2023llama}, and Llama-2-7B with instruction-tuning were used. We followed the prompts and instruction-tuning approaches published by \cite{vats-etal-2023-llms} in a few-shot setup, and used the PredEx training data for instruction-tuning.

\subsubsection{Prompts Used}
For inference, we utilized prompts published by \cite{vats-etal-2023-llms} and employed Template 2 in a zero and few-shot setup exclusively for prediction tasks, as detailed in Table \ref{tab:judgment_prediction_prompts_few} in the Appendix. These prompts provide a case description alongside a gold standard prediction label, directing the LLM to generate the judicial decision. For instruction tuning, we employed our custom prompts for prediction tasks, which are listed in Table \ref{tab:judgment_prediction_prompts_zero} in the Appendix of this paper.

\subsubsection{Instruction-Set}
We developed 16 instruction sets using ChatGPT4 (DALL-E), validated by legal experts and then used for PredEx training data for instruction tuning. Given the token limit of 4096 in LLMs, we selected the last 1000 words from each document to fit within this constraint. This choice is supported by findings from \cite{malik-etal-2021-ildc} who achieved optimal results using the last 512 tokens of judgments. The input comprised the case proceedings and case decision and a random selection of instructions, with the output being the case outcome prediction. For a comprehensive understanding of our methodology and the full range of instructions used, we have included the complete list of all 16 instruction sets in Table~\ref{Instruction-sets} located in the Appendix of this paper.

\subsection{Judgment Prediction with Explanation}
For this task, we employed the same LLMs with settings similar to the Judgment Prediction task, but with different instructions focusing on both prediction and explanation.

\subsubsection{Prompts used}
In our approach, we also adopted prompts from \cite{vats-etal-2023-llms}, specifically utilizing Template 1 for the combined task of judgment prediction and explanation. This template, detailed in Table \ref{tab:judgment_prediction_prompts_few} in the Appendix, expands on the prediction-only format of Template 2 by including an explanation component. Here also, for instruction tuning, we employed our custom prompts for prediction with explanation tasks, which are listed in Table \ref{tab:judgment_prediction_prompts_zero} in the Appendix of this paper. In this setup, the LLM is instructed not only to predict the outcome of a case but also to articulate the reasoning behind the decision. The precise formatting and examples of how predictions and explanations are structured and solicited from the LLM can be viewed in the referenced table.

\subsubsection{Instruction-Set}
For judgment prediction with explanation, we created 16 instruction sets using ChatGPT4 (DALL-E), also validated by legal experts. This time, the input included case proceedings, decisions, and reasoning, with randomly chosen instructions, and the output being the case outcome prediction with explanation. For a comprehensive view of all 16 instruction sets, we have included the full list in Table~\ref{Instruction-sets} in the Appendix of this paper.
