\section{Related Work}
The field of Legal Judgment Prediction (LJP) has seen significant advancements, driven by the need to automate legal case outcome forecasting and alleviate the burden of overwhelming caseloads. Early works by \cite{aletras2016predicting}, \cite{chalkidis2019neural}, and \cite{feng2021recommending} laid the foundation for LJP, emphasizing the importance of explainability in AI predictions. Benchmark datasets such as CAIL2018 \cite{xiao2018cail2018}, ECHR-CASES \cite{chalkidis2019neural}, and others have spurred research in this area, inspiring models like TopJudge and MLCP-NLN. However, there remains a gap between machine and human performance.

In the Indian context, datasets like ILDC \cite{malik-etal-2021-ildc}, PredEx \cite{nigam-etal-2024-legal} and \cite{nigam2022nigam, malik-etal-2022-semantic, nigam2023legal} have highlighted the growing role of AI in legal judgments, with an emphasis on explainability. Research in LJP with LLMs, such as \cite{vats2023llms} and \cite{nigam-etal-2024-legal}, has experimented with models like GPT-3.5 Turbo and Llama-2 on Indian legal datasets. Other studies, such as \cite{masala2021jurbert} on Romanian legal texts and \cite{hwang2022multi} on Korean legal language, have demonstrated LJP's adaptability across legal systems.

Cross-jurisdictional work, including \cite{zhao-etal-2018-learning}, showcases LJP's applicability in different legal frameworks, with research expanding to multilingual considerations, as seen in \cite{niklaus2021swiss} and \cite{kapoor-etal-2022-hldc} for Hindi legal documents. Recent innovations, such as event extraction and multi-stage learning~\cite{feng-etal-2022-legal}, continue to push the boundaries of LJP research.

% In summary, the LJP field is dynamic and multifaceted, encompassing both practical applications and theoretical inquiries into AI explainability. Our work seeks to contribute to this evolving landscape by developing a comprehensive English legal dataset for India and investigating advanced models capable of processing complete legal judgments in realistic scenarios.







% Legal judgment Prediction~\cite{strickson2020legal,xu2020multi,feng2023criminal} is the task of predicting decisions of legal judgments. The task of judgment Prediction~\cite{xu2020multi} has been performed over several countries which use a variety of machine learning models. The task of Legal judgment Prediction~\cite{malik-etal-2021-ildc,nigam2023nonet,nigam-etal-2024-legal} has been attempted in the Indian Context by creating a legal judgment dataset and performing the task of judgment prediction using hierarchical transformer models along with providing an explanation for the judgment. Legal judgment Prediction with facts~\cite{nigam2023fact} has been attempted in which they run some transformer models only on the case facts of the ILDC corpus.
% Legal judgment Prediction with LLMs~\cite{vats2023llms,nigam-etal-2024-legal} has been tried on the ILDC and PredEx corpus where they experiment with several LLMs. Legal judgment prediction~\cite{masala2021jurbert} with the Romanian BERT model has been tried. Legal judgment Prediction~\cite{hwang2022multi} with a multi-task benchmark for Korean legal language
% understanding and judgment prediction have been tried.